An indexed construction of semi-simplicial and semi-cubical types

### 2022-05-16 paper/paper.tex
README: update loc; it's ~800
paper: introduce magic joinargs lua code!

% Universes
\newcommand{\sort}[1]{\ensuremath{\mathsf{U}_{#1}}}

@@

% Restriction for frame, layer, and filler
\newcommand{\restrf}[2]{\ensuremath{\mathsf{restr}_{\mathsf{frame}, #1}^{#2}}}
\newcommand{\restrl}[2]{\ensuremath{\mathsf{restr}_{\mathsf{layer}, #1}^{#2}}}
\newcommand{\restrc}[2]{\ensuremath{\mathsf{restr}_{\mathsf{filler}, #1}^{#2}}}
\newcommand{\fullframe}[2]{\ensuremath{\mathsf{fullframe}_{#1}^{#2}}}

@@

$$
  \begin{tikzcd}
    X_0 : \sort{} & X_1 : \sort{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \sort{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
$$

\begin{align*}
  X_0 & : \sort{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \sort{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \sort{} \\
  \ldots
\end{align*}

@@

\subsection{Truncated sets}
The definition is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. It describes the structure of the underlying higher-dimensional relations on which cubes and boxes are built, together with the definition of homogeneous $n$-boxes, homogeneous $n$-cubes, together with face operations on boxes and cubes, together with commutation properties of the face operations. All are mutually defined as types of the target language ETT. Note that such relational structure, cubes and boxes are relative to a universe. Note the presence of a coherence condition $\cohframe{l}{}$ to ensuring that both sides of the equality in $\restrl{l}{}$ and $\restrc{l}{}$ are in the same type. The proof of $\cohframe{l}{}$ itself requires an higher-dimensional coherence condition which we obtain by working here in ETT where all proofs of an equality are identified (principle of Unicity of Identity Proofs). Note that if the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory). Note also that for a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf{l}{}$ nor $\cohframe{l}{}$ anymore.

@@

\begin{eqntable}{Truncated sets, the core\label{tab:core}}
  \mc{Type of truncated sets}
  \eqnline{\X{l}{n}}{}{:}{\sort{l+1}}
  \eqnline{\X{l}{0}}{}{\defeq}{\unittype}
  \eqnline{\X{l}{n'+1}}{}{\defeq}{\ensuremath{\Sigma}D:\X{l}{n'}.\,\Xcomp{l}{n'} \D}

@@

  \mc{Structure carried at each dimension}
  \eqnline{\Xcomp{l}{n}}{(D:\X{l}{n})}{:}{\sort{l}}
  \eqnline{\Xcomp{l}{n}}{D}{\defeq}{\fullframe{l}{n}\D\imp \sort{l}}
\end{eqntable}

\begin{eqntable}{$\mathsf{frame}, \mathsf{layer}, \mathsf{filler}$\label{tab:frames}}
  \mc{Full $n$-frames}
  \eqnline{\fullframe{l}{n}}{(D:\X{l}{n})}{:}{\sort{l}}
  \eqnline{\fullframe{l}{n}}{D}{\defeq}{\framep{l}{n,n}\D}

@@

  \mc{Partial $n$-frames, $n$-layers, and $n$-fillers}
  \eqnline{\framep{l}{n,p,[p \leq n]}}{(D:\X{l}{n})}{:}{\sort{l}}
  \eqnline{\framep{l}{n,0}}{D}{\defeq}{\unittype}
  \eqnline{\framep{l}{n,p'+1}}{D}{\defeq}{\Sigma d:\framep{l}{n,p'}\D.\,\layer{l}{n,p'} \D \d}

@@

  \eqnline{\layer{l}{n,p,[p < n]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\sort{l}}

  \eqnline{\layer{l}{n,p}}{D~d}{\defeq}{\makecell{\filler{l}{n-1,p}\hdD\tlD(\restrf{l,L,p}{n,p} \D \d) \\ \times\;\filler{l}{n-1,p}\hdD\tlD(\restrf{l,R,p}{n,p} \D \d)}}

@@

  \eqnline{\filler{l}{n,p,[p \leq n]}}{\makecell{(D:\X{l}{n}) \\(E:\Xcomp{l}{n} \D) \\ (d:\framep{l}{n,p} \D)}}{:}{\sort{l}}

@@

\begin{eqntable}{$q$-th projection of $\mathsf{restr}$, or faces\label{tab:faces}}
  \eqnline{\restrf{l,\epsilon,q}{n,p,[p \leq q \leq n - 1]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\framep{l}{n-1,p}\hdD}

  \eqnline{\restrf{l,\epsilon,q}{n,p'+1}}{D~(d,l)}{\defeq}{(\restrf{l,\epsilon,q}{n,p'} \D\d,\restrl{l,\epsilon,q-1}{n,p'} \D \d\l)}

@@

  \eqnline{\restrl{l,\epsilon,q}{n,p,[p \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D) \\ (b:\layer{l}{n,p} \D \d)}}{:}{\layer{l}{n-1,p}\hdD(\restrf{l,\epsilon,q+1}{n,p} \D \d)}

  \eqnline{\restrl{l,\epsilon,q}{n,p}}{D~d~c}{\defeq}{\makecell{(\overright{\cohframe{l,\epsilon,L,q,p}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf{l,L,p}{n,p} \D \d)(c_L)), \\ \;\overright{\cohframe{l,\epsilon,R,q,p}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf{l,R,p}{n,p}\D\d)(c_R)))}}

@@

  \eqnline{\restrc{l,\epsilon,q}{n,p,[p \leq q \leq n - 1]}}{\makecell{(D:\X{l}{n}) \\ (E:\Xcomp{l}{n} \D) \\(d:\framep{l}{n,p} \D) \\ (b:\filler{l}{n,p} \D\E\d)}}{:}{\filler{l}{n-1,p}\hdD\tlD(\restrf{l,\epsilon,q+1}{n,p} \D \d)}

@@

  \eqnline{\cohframe{l,\epsilon,\omega,q,r}{n,p,[p \leq r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\makecell{\restrf{l,\epsilon,q}{n-1,p}\hdD(\restrf{l,\omega,r}{n,p} \D \d) \\ \eqett \restrf{l,\omega,r}{n-1,p}\hdD(\restrf{l,\epsilon,q+1}{n,p} \D \d)}}

@@

  \eqnline{\cohlayer{l,\epsilon,\omega,q,r}{n,p,[p < r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D) \\(b:\layer{l}{n,p} \D \d)}}{:}{\makecell{\overright{\cohframe{l,\epsilon,\omega,q+1,r+1}{n,p} \D \d}(\restrl{l,\epsilon,q}{n-1,p}\hdD(\restrf{l,\omega,r}{n,p} \D \d)(\restrl{l,\omega,r}{n,p} \D \d(b))) \\ \eqett \restrl{l,\omega,r}{n-1,p}\hdD(\restrf{l,\epsilon,q+1}{n,p} \D \d)(\restrl{l,\epsilon,q+1}{n,p} \D \d(b))}}

  \eqnline{\cohlayer{l,\epsilon,\omega,q,r}{n,p}}{D~d~c}{\defeq}{\makecell{(\cohfiller{l,\epsilon,\omega,q-1,r-1}{n-1,p}\hdD\tlD(\restrf{l,L,p}{n,p} \D \d)(c_L), \\ \;\cohfiller{l,\epsilon,\omega,q-1,r-1}{n-1,p}\hdD\tlD(\restrf{l,R,p}{n,p} \D \d)(c_R))}}

@@

  \eqnline{\cohfiller{l,\epsilon,\omega,q,r}{n,p,[p \leq r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (E:\Xcomp{l}{n} \D) \\ (d:\framep{l}{n,p} \D) \\ (b:\filler{l}{n,p} \D\E \d)}}{:}{\makecell{\overright{\cohframe{l,\epsilon,\omega,q+1,r+1}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf{l,\omega,r}{n,p} \D \d)(\restrc{l,\omega,r}{n,p} \D\E \d(b))) \\ \eqett \restrc{l,\omega,r}{n-1,p}\hdD\tlD(\restrf{l,\epsilon,q+1}{n,p} \D \d)(\restrc{l,\epsilon,q+1}{n,p} \D\E \d(b))}}

  \eqnline{\cohfiller{l,\epsilon,\omega,q,r}{n,p,[p=r]}}{D~E~d~(b,\_)}{\defeq}{\reflett(\restrc{l,\epsilon,q-1}{n-1,p}\hdD\tlD(\restrf{l,\omega,p}{n,p} \D \d)(b_{\epsilon}))}

@@

\begin{eqntable}{Coinductive structure\label{tab:coind}}
  \eqnline{\Xp{l}}{}{:}{\sort{l+1}}
  \eqnline{\Xp{l}}{}{\defeq}{\Xfrom{l}{0}(\star)}
  \\
  \eqnline{\Xfrom{l}{n}}{(D:\X{l}{n})}{:}{\sort{l+1}}
  \eqnline{\Xfrom{l}{n}}{D}{\defeq}{\Sigma R:\Xcomp{l}{n} \D.\,\Xfrom{l}{n+1}(D,R)}
\end{eqntable}

>>>>

% Universes
\newcommand{\U}[1]{\ensuremath{\mathsf{U}_{#1}}}

@@

% Comma-separeated optional arguments
\begin{luacode*}
  function remove_empty(tbl)
    for i, v in ipairs(tbl) do
        if v == "" then
          table.remove(tbl, i)
          return remove_empty(tbl)
        end
    end
    return tbl
  end
  function join_with_comma(tbl)
    return table.concat(remove_empty(tbl), ",")
  end
\end{luacode*}

\newcommandx{\joinargs}[4][1=,2=,3=,4=]{\directlua{tex.sprint(join_with_comma({"\luaescapestring{#1}";"\luaescapestring{#2}";"\luaescapestring{#3}";"\luaescapestring{#4}"}))}}

% Restriction for frame, layer, and filler
\newcommandx{\restrf}[6][1=,2=,3=,4=,5=,6=]{\ensuremath{\mathsf{restr}_{\mathsf{frame}, \joinargs[#1][#2][#3]}^{\joinargs[#4][#5][#6]}}}
\newcommand{\restrl}[2]{\ensuremath{\mathsf{restr}_{\mathsf{layer}, #1}^{#2}}}
\newcommand{\restrc}[2]{\ensuremath{\mathsf{restr}_{\mathsf{filler}, #1}^{#2}}}
\newcommand{\fullframe}[2]{\ensuremath{\mathsf{fullframe}_{#1}^{#2}}}

@@

$$
  \begin{tikzcd}
    X_0 : \U{} & X_1 : \U{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
$$

\begin{align*}
  X_0 & : \U{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \U{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \U{} \\
  \ldots
\end{align*}

@@

\subsection{Truncated sets}
The definition is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. It describes the structure of the underlying higher-dimensional relations on which cubes and boxes are built, together with the definition of homogeneous $n$-boxes, homogeneous $n$-cubes, together with face operations on boxes and cubes, together with commutation properties of the face operations. All are mutually defined as types of the target language ETT. Note that such relational structure, cubes and boxes are relative to a universe. Note the presence of a coherence condition $\cohframe{l}{}$ to ensuring that both sides of the equality in $\restrl{l}{}$ and $\restrc{l}{}$ are in the same type. The proof of $\cohframe{l}{}$ itself requires an higher-dimensional coherence condition which we obtain by working here in ETT where all proofs of an equality are identified (principle of Unicity of Identity Proofs). Note that if the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory). Note also that for a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf[l]$ nor $\cohframe{l}{}$ anymore.

@@

\begin{eqntable}{Truncated sets, the core\label{tab:core}}
  \mc{Type of truncated sets}
  \eqnline{\X{l}{n}}{}{:}{\U{l+1}}
  \eqnline{\X{l}{0}}{}{\defeq}{\unittype}
  \eqnline{\X{l}{n'+1}}{}{\defeq}{\ensuremath{\Sigma}D:\X{l}{n'}.\,\Xcomp{l}{n'} \D}

@@

  \mc{Structure carried at each dimension}
  \eqnline{\Xcomp{l}{n}}{(D:\X{l}{n})}{:}{\U{l}}
  \eqnline{\Xcomp{l}{n}}{D}{\defeq}{\fullframe{l}{n}\D\imp \U{l}}
\end{eqntable}

\begin{eqntable}{frame, layer, and filler\label{tab:frames}}
  \mc{Full $n$-frames}
  \eqnline{\fullframe{l}{n}}{(D:\X{l}{n})}{:}{\U{l}}
  \eqnline{\fullframe{l}{n}}{D}{\defeq}{\framep{l}{n,n}\D}

@@

  \mc{Partial $n$-frames, $n$-layers, and $n$-fillers}
  \eqnline{\framep{l}{n,p,[p \leq n]}}{(D:\X{l}{n})}{:}{\U{l}}
  \eqnline{\framep{l}{n,0}}{D}{\defeq}{\unittype}
  \eqnline{\framep{l}{n,p'+1}}{D}{\defeq}{\Sigma d:\framep{l}{n,p'}\D.\,\layer{l}{n,p'} \D \d}

@@

  \eqnline{\layer{l}{n,p,[p < n]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\U{l}}

  \eqnline{\layer{l}{n,p}}{D~d}{\defeq}{\makecell{\filler{l}{n-1,p}\hdD\tlD(\restrf[l][L][p][n][p] \D \d) \\ \times\;\filler{l}{n-1,p}\hdD\tlD(\restrf[l][R][p][n][p] \D \d)}}

@@

  \eqnline{\filler{l}{n,p,[p \leq n]}}{\makecell{(D:\X{l}{n}) \\(E:\Xcomp{l}{n} \D) \\ (d:\framep{l}{n,p} \D)}}{:}{\U{l}}

@@

\begin{eqntable}{$q$-th projection of $\mathsf{restr}$, or faces\label{tab:faces}}
  \eqnline{\restrf[l][\epsilon][q][n][p][p \leq q \leq n - 1]}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\framep{l}{n-1,p}\hdD}

  \eqnline{\restrf[l][\epsilon][q][n][p'+1]}{D~(d,l)}{\defeq}{(\restrf[l][\epsilon][q][n][p'] \D\d,\restrl{l,\epsilon,q-1}{n,p'} \D \d\l)}

@@

  \eqnline{\restrl{l,\epsilon,q}{n,p,[p \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D) \\ (b:\layer{l}{n,p} \D \d)}}{:}{\layer{l}{n-1,p}\hdD(\restrf[l][\epsilon][q+1][n][p] \D \d)}

  \eqnline{\restrl{l,\epsilon,q}{n,p}}{D~d~c}{\defeq}{\makecell{(\overright{\cohframe{l,\epsilon,L,q,p}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf[l][L][p][n][p] \D \d)(c_L)), \\ \;\overright{\cohframe{l,\epsilon,R,q,p}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf[l][R][p][n][p] \D\d)(c_R)))}}

@@

  \eqnline{\restrc{l,\epsilon,q}{n,p,[p \leq q \leq n - 1]}}{\makecell{(D:\X{l}{n}) \\ (E:\Xcomp{l}{n} \D) \\(d:\framep{l}{n,p} \D) \\ (b:\filler{l}{n,p} \D\E\d)}}{:}{\filler{l}{n-1,p}\hdD\tlD(\restrf[l][\epsilon][q+1][n][p] \D \d)}

@@

  \eqnline{\cohframe{l,\epsilon,\omega,q,r}{n,p,[p \leq r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D)}}{:}{\makecell{\restrf[l][\epsilon][q][n-1][p]\hdD(\restrf[l][\omega][r][n][p] \D \d) \\ \eqett \restrf{l,\omega,r}{n-1,p}\hdD(\restrf[l][\epsilon][q+1][n][p] \D \d)}}

@@

  \eqnline{\cohlayer{l,\epsilon,\omega,q,r}{n,p,[p < r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (d:\framep{l}{n,p} \D) \\(b:\layer{l}{n,p} \D \d)}}{:}{\makecell{\overright{\cohframe{l,\epsilon,\omega,q+1,r+1}{n,p} \D \d}(\restrl{l,\epsilon,q}{n-1,p}\hdD(\restrf[l][\omega][r][n][p] \D \d)(\restrl{l,\omega,r}{n,p} \D \d(b))) \\ \eqett \restrl{l,\omega,r}{n-1,p}\hdD(\restrf[l][\epsilon][q+1][n][p] \D \d)(\restrl{l,\epsilon,q+1}{n,p} \D \d(b))}}

  \eqnline{\cohlayer{l,\epsilon,\omega,q,r}{n,p}}{D~d~c}{\defeq}{\makecell{(\cohfiller{l,\epsilon,\omega,q-1,r-1}{n-1,p}\hdD\tlD(\restrf[l][L][p][n][p] \D \d)(c_L), \\ \;\cohfiller{l,\epsilon,\omega,q-1,r-1}{n-1,p}\hdD\tlD(\restrf[l][R][p][n][p] \D \d)(c_R))}}

@@

  \eqnline{\cohfiller{l,\epsilon,\omega,q,r}{n,p,[p \leq r \leq q \leq n - 2]}}{\makecell{(D:\X{l}{n}) \\ (E:\Xcomp{l}{n} \D) \\ (d:\framep{l}{n,p} \D) \\ (b:\filler{l}{n,p} \D\E \d)}}{:}{\makecell{\overright{\cohframe{l,\epsilon,\omega,q+1,r+1}{n,p} \D \d}(\restrc{l,\epsilon,q}{n-1,p}\hdD\tlD(\restrf[l][\omega][r][n][p] \D \d)(\restrc{l,\omega,r}{n,p} \D\E \d(b))) \\ \eqett \restrc{l,\omega,r}{n-1,p}\hdD\tlD(\restrf[l][\epsilon][q+1][n][p] \D \d)(\restrc{l,\epsilon,q+1}{n,p} \D\E \d(b))}}

  \eqnline{\cohfiller{l,\epsilon,\omega,q,r}{n,p,[p=r]}}{D~E~d~(b,\_)}{\defeq}{\reflett(\restrc{l,\epsilon,q-1}{n-1,p}\hdD\tlD(\restrf[l][\omega][p][n][p] \D \d)(b_{\epsilon}))}

@@

\begin{eqntable}{Coinductive structure\label{tab:coind}}
  \eqnline{\Xp{l}}{}{:}{\U{l+1}}
  \eqnline{\Xp{l}}{}{\defeq}{\Xfrom{l}{0}(\star)}
  \\
  \eqnline{\Xfrom{l}{n}}{(D:\X{l}{n})}{:}{\U{l+1}}
  \eqnline{\Xfrom{l}{n}}{D}{\defeq}{\Sigma R:\Xcomp{l}{n} \D.\,\Xfrom{l}{n+1}(D,R)}
\end{eqntable}
